Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1)

### 2014-04-25 WSSSPE1_report/WSSSPE1_summary_report.tex
(1) Shuffle Appendix A and Appendix B and change related mentions. (2) Move Table 1 from section 7 to section 5 where it is first mentioned. (3) Minor changes to Section 5.

This report attempts to summarize the various aspects of the workshop.
The remainder of this paper first gives an overview of the process
with which the workshop was put together (\S\ref{sec:process}), then
proceeds with summaries of the two keynote
presentations~(\S\ref{sec:keynotes}), followed by summaries of the
workshop papers grouped by the three thematic workshop panels to which
they were assigned~(\S\ref{sec:devel}-\ref{sec:community}). Three
broader cross-cutting issues surfaced repeatedly, and are discussed
separately~(\S\ref{sec:cross-cutting}), as are use cases for
sustainable software~(\S\ref{sec:use-cases}). The summaries are based
not only on the papers and panel presentations, but also on the many
comments raised in both the onsite and online discussions that
accompanied the workshop elements, as documented by collaborative note
taking during the workshop~\cite{WSSSPE1-google-notes}. We conclude
with issues and lessons learned, as well as plans for future
activities~(\S\ref{sec:conclusions}).  The short papers accepted to
the workshop are listed in Appendix~\ref{sec:papers}, part of he original
call for papers is included in 
Appendix~\ref{sec:cfp}, and a partial
list of workshop attendees can be found in
Appendix~\ref{sec:attendees}.

@@

The workshop submissions grouped under this section provide frameworks
for thinking about, researching and understanding which elements of
sustainability are important and how they relate. Although there is
substantial overlap between the frameworks, they have different
emphases and extents.  Each paper in this group included a definition
of sustainability, with many overlaps between them (see
Table~\ref{tab:defining-sustainability}).  Perhaps unsurprisingly, the
issue of how to define sustainability came to the fore multiple times
during the workshop, and it is thus summarized in depth separately in
~\S\ref{sec:defining-sustainability}.

@@

One group of papers presented frameworks that were primarily about
characteristics of software artifacts, connecting with the long
discourse on software quality. This approach is realized in adjectives
that can be applied to pieces of software but might also extend to
describe software projects.  Thus Calero et al.~\cite{Calero_WSSSPE}
propose adding elements to the ISO standards for measuring software
quality. Specifically, they propose an additional dimension of quality
they call ``perdurability'' or ability to ``last over time'' made up
of three characteristics: reliability, maintainability and
adaptability. This shows some overlap with the framework employed by
Venters et al.~\cite{Venters_WSSSPE} who employ the adjectives of
``extensibility, interoperability, maintainability, portability,
reusability and scalability,'' anticipating the sorts of work that
people would need to do with a software artifact in the future, ``as
stakeholders requirements, technology and environments evolve and
change.'' Venters et al.~argue that these choices are really made very
early because they relate to the architecture of the software and
involve trade offs that ought to be analyzed alongside other
attributes of the software artifact. Lenhardt et
al.~\cite{Lenhardt_WSSSPE} compare the software lifecycle to the data
lifecycle to argue for the inclusion of metadata throughout a piece of
software's life (discussing, for example, how it has been built and
tested and what ``data in'' and ``data out'' has been considered). In
addition, their analogy suggests that the software lifecycle might add
a phase of ``preservation'' and draw on understandings of what that
involves from studies of data.  In sum, then, these frameworks focused
on what needs to be accomplished to have more sustainable software.

A second theme in these papers was the continued availability of
resources to accomplish the goals of sustainability. The elements of
these frameworks focused far more on the organization of a software
project than they did on characteristics of the artifact itself
(although it is certainly true that the adjectives discussed above
could be applied to a software project). For example Pierce et
al.~\cite{Pierce_WSSSPE} focus on the way that the project is run,
particularly in terms of how those involved communicate and jointly
set priorities, a process they call governance. In particular they
argue that because sustainability relates to having ongoing resources,
governance must be open to receive diverse input (by occurring online,
asynchronously, and publicly) and thus have the potential to
``transform passive users into active contributors''. They argue that
the Apache Software Foundation's incubation process teaches this and
could be learned from by projects throughout scientific software.
Katz and Proctor~\cite{Katz_WSSSPE} also discuss governance,
describing two modes: ``top-down'' and ``bottom-up'' governance.  They
place governance alongside questions of the artifact (``technical''),
and questions of the origins of resources, in terms of who is funding
the work surrounding a piece of software (``the political'') and the
manner in which resources come to the project initially (commercial,
open source, closed partnerships, grant funded) and long-term (all
four plus paid support).

@@

\appendix
\section{Papers Accepted and Discussed at WSSSPE1} \label{sec:papers}

>>>>

This report attempts to summarize the various aspects of the workshop.
The remainder of this paper first gives an overview of the process
with which the workshop was put together (\S\ref{sec:process}), then
proceeds with summaries of the two keynote
presentations~(\S\ref{sec:keynotes}), followed by summaries of the
workshop papers grouped by the three thematic workshop panels to which
they were assigned~(\S\ref{sec:devel}-\ref{sec:community}). Three
broader cross-cutting issues surfaced repeatedly, and are discussed
separately~(\S\ref{sec:cross-cutting}), as are use cases for
sustainable software~(\S\ref{sec:use-cases}). The summaries are based
not only on the papers and panel presentations, but also on the many
comments raised in both the onsite and online discussions that
accompanied the workshop elements, as documented by collaborative note
taking during the workshop~\cite{WSSSPE1-google-notes}. We conclude
with issues and lessons learned, as well as plans for future
activities~(\S\ref{sec:conclusions}).  The original call for papers is included in 
Appendix~\ref{sec:cfp}.  The short papers accepted to
the workshop are listed in Appendix~\ref{sec:papers}, and a partial
list of workshop attendees can be found in
Appendix~\ref{sec:attendees}.

@@

The workshop submissions grouped under this section provide frameworks
for thinking about, researching, and understanding which elements of
sustainability are important and how they are related to each other. Although there is
substantial overlap between the frameworks, they have different
emphases and extents.  Each paper in this group included a definition
of sustainability, with many overlaps between them (see
Table~\ref{tab:defining-sustainability}).  Perhaps unsurprisingly, the
issue of how to define sustainability came to the fore multiple times
during the workshop, and it is thus summarized in depth separately in
~\S\ref{sec:defining-sustainability}.

@@

One group of papers presented frameworks that were primarily about
characteristics of software artifacts, connecting with the long
discourse on software quality. This approach is realized in adjectives
that can be applied to pieces of software but might also extend to
describe software projects.  Thus Calero et al.~\cite{Calero_WSSSPE}
propose adding elements to the ISO standards for measuring software
quality. Specifically, they propose an additional dimension of quality
they call ``perdurability'' with
three defining characteristics: reliability, maintainability, and
adaptability. This overlaps with the framework   by
Venters et al.~\cite{Venters_WSSSPE} who employ the features
``extensibility, interoperability, maintainability, portability,
reusability and scalability,'' anticipating the sorts of work that
people would need to do with a software artifact in the future, ``as
stakeholders requirements, technology and environments evolve and
change.'' Venters et al.~argue that these choices need to be made 
early because they are related to the architecture of the software and
involve trade offs that ought to be analyzed alongside each other. Lenhardt et
al.~\cite{Lenhardt_WSSSPE} compare the software lifecycle to the data
lifecycle to argue for the inclusion of metadata throughout a piece of
software's life (discussing, for example, how it has been built and
tested and what ``data in'' and ``data out'' has been considered). In
addition, their analogy suggests that the software lifecycle might add
a phase of ``preservation'' and draw on the understanding of what that
involves from studies of data.  In sum, then, these frameworks focused
on what needs to be accomplished to have more sustainable software.

A second theme in these papers was the continued availability of
resources to accomplish the goals of sustainability. The elements of
these frameworks focused far more on the organization of a software
project than they did on characteristics of the artifact itself
(although it is certainly true that the adjectives discussed above
could be applied to a software project). For example Pierce et
al.~\cite{Pierce_WSSSPE} focus on the way that the project is run,
particularly in terms of how those involved communicate and jointly
set priorities, a process they call governance. In particular they
argue that because sustainability is related to having ongoing resources,
governance must be open to receive diverse input (by occurring online,
asynchronously, or publicly) and thus have the potential to
``transform passive users into active contributors.'' They argue that
the Apache Software Foundation's incubation process teaches this and
could be learned from by projects throughout scientific software.
Katz and Proctor~\cite{Katz_WSSSPE} also discuss governance,
describing two modes: ``top-down'' and ``bottom-up'' governance.  They
place governance alongside questions of the artifact (``technical''),
and questions of the origins of resources, in terms of who is funding
the work surrounding a piece of software (``the political'') and the
manner in which resources come to the project initially (commercial,
open source, closed partnerships, grant funded) and long-term (all
four plus paid support).

@@

\appendix

\section{Papers Accepted and Discussed at WSSSPE1} \label{sec:papers}
